Blum axioms and nondeterministic computation of functions

### 2016-04-20 paper.tex
paper.tex: Improve aesthetics.

In Theory of Computation,
we usually use languages
to mathematically model problems in the real world.
Decision problems (``yes/no'') are mapped to languages in a very natural way,
by just putting every ``yes'' instance in the language,
and leaving the rest out.
Search problems usually are rewritten as a decision problem,
and then this problem is converted to a language.
For instance,
the task of finding a satisfying assignment for a Boolean formula
is reinterpreted to the task of deciding whether this assignment exists,
and this task is then converted to a language
--- in this example we have $\SAT$, the Boolean satisfiability problem.
This does the trick when it comes to proving that something is hard;
for instance, if we show that the decision problem is $\NP$-hard or undecidable,
then intuitively the corresponding search problem must be at least as hard.
Therefore,
concepts like ``decidable'', ``$\NP$-complete'', ``polynomial-time decidable''
arise naturally in the context of decision problems.

However, even from the theoretical standpoint,
it is useful to extend these concepts to functions.
For instance,
the concept of polynomial-time computable functions
are required to define Karp reductions \cite[p.~42]{AroraBarak2009}.
For single-tape deterministic Turing machines,
this definition is easy to extend:
A function $f : \{0, 1\}^* \to \{0, 1\}^*$
is said to be \emph{polynomial-time computable}
if there is a Turing machine $M$ and a polynomial $p$ such that,
when given the string $x$ as input,
$M$ halts with $f(x)$ in its tape within $p(|x|)$ steps.

@@

This paper presents several attempts to establish
how a nondeterministic Turing machine could compute a function,
and to extend the concept of ``polynomial-time computable'' under each definition.
Section~\ref{sec:quality-standards} sets reasonableness criteria
to both the definition of function computation and complexity of the computation.
Section~\ref{sec:nondeterministic_computation_of_functions},
which contains the several definitions,
presents and criticizes Hopcroft and Ullman's and Goldreich's definitions
(sections \ref{sec:hopcroft-ullman} and~\ref{sec:goldreich}),
proposes one possible definition that meet the quality standards
(section~\ref{sec:proposed}),
and shows other two definitions,
by Krentel (sec~\ref{sec:krentel}) and Valiant (sec~\ref{sec:valiant}),
that, altough were not proposed in the context
of general nondeterministic computation of functions,
also meet the quality standards and sidesteps the problems with the proposed definition
(mentioned in section~\ref{sec:np-completeness-extension}).
The development of this project is given in section~\ref{sec:development},
after the concluding remarks (section~\ref{sec:conclusion}).

@@

One important feature about the standard encodings of deterministic Turing machines
is the Universal Turing Machine Theorem:
the existence of a partial recursive function
$U: \{0, 1\}^* \times \{0, 1\}^* \to \{0, 1\}^*$
% Não sei se $U$ é um bom nome...
such that, if $w$ encodes the machine $M$,
then $U(m, x)$ is the result of running the machine $M$ on $x$.
The machines that compute $U$ are called \emph{universal Turing machines}.

@@

\begin{definition}
    Let $\mathcal P$ be the set of all partial recursive functions.
    An \emph{acceptable Gödel numbering}
    is a function $\phi : \{0, 1\}^* \to \mathcal P$,
    that associates each string (or program)\footnotemark $w \in \{0, 1\}^*$
    \footnotetext{
        In texts like Rogers'~\cite{Rogers1987},
        the partial recursive functions have the naturals as domain and codomain
        (that is, they are of the form $f: \mathbb N \to \mathbb N$),
        and Gödel numberings associates natural numbers with recursive functions.
        In this paper we will work with binary strings instead of numbers,
        which will simplify the definition of complexity classes
        and allows us to think the string $w$ as a \emph{program} for $\phi_w$
        (see example~\ref{c}).
    }
    to a function $\phi_w \in \mathcal P$,
    that satisfies
    \begin{enumerate}
        \item $\phi$ is surjective;
            that is, every partial recursive function $f \in \mathcal P$
            has a program $w \in \{0, 1\}^*$ such that $\phi_w = f$;
            \label{surjectiveness}
        \item There is a partial recursive function
            $U:\{0, 1\}^* \times \{0, 1\}^* \to \{0, 1\}^*$ such that,
            for every $w$ and $x$,
            $U(w, x)$ is defined if and only if $\phi_w(x)$ is defined,
            and, in this case,
            \begin{equation*}
                U(w, x) = \phi_w(x);
            \end{equation*}
            \label{universal-tm}
        \item There is a total recursive function
            $\sigma:\{0, 1\}^* \times \{0, 1\}^* \to \{0, 1\}^*$ such that,
            for every $y$,
            $\phi_w(x, y)$ is defined if and only if $\phi_{\sigma(w, x)}(y)$ is defined,
            and, in this case,
            \begin{equation*}
                \phi_w(x, y) = \phi_{\sigma(w, x)}(y).
            \end{equation*}
            \label{smn-theorem}
    \end{enumerate}
\end{definition}

@@

Condition~\ref{smn-theorem} is the ``little more'' we mentioned earlier.
It is known as the $S_{mn}$ theorem \cite[p.~24]{Rogers1987}.
Essentially,
given a partial recursive function of two variables,
we can obtain a partial recursive function of one variable
by fixing the first argument.
The function $\sigma$ provides a systematic way of doing this:
given a description $w$ of a partial recursive function of two variables
and the value $x$ to be fixed as the first variable,
the function returns a machine $\sigma(w, x)$
that computes the same function as $w$ but with the first argument fixed as $x$.

\begin{example}\label{standard-numbering}
    Any encoding of deterministic Turing machines as a binary string
    yields an acceptable numbering of the recursive functions.
\end{example}

\begin{example}\label{c}
    Any programming language can be understood as an acceptable Gödel numbering.
    For example,
    if we restrict a \texttt C program
    to perform input and output only using the standard input and standard output
    (that is, forbid interactions with the user, file reading, GUIs,
    access to system clock, etc.),
    the resulting program will map a binary input to a binary output,
    characterizing a partial recursive function.
    Thus, we can see a \texttt C compiler
    as an implementation of an acceptable Gödel numbering.\footnote{
        Note we are ignoring here issues like compilation and run-time errors.
        These can be dealt with as in the case of Turing machines:
        any invalid program will signify some fixed partial recursive function
        (say, the function that is defined nowhere);
        and any invalid step in computation
        makes the function to be not defined on that input.
    }
\end{example}

@@

\begin{definition}
    Given an acceptable Gödel numbering $\phi$,
    a \emph{complexity measure} for $\phi$
    is a function $\Phi:\Sigma^* \times \Sigma^* \to \mathbb N$ of two variables
    that satisfies: \cite[p.~324]{Blum1967}
    \begin{enumerate}
        \item For every $w$ and $x$,
            $\phi_w(x)$ exists if and only if $\Phi(w, x)$ exists.
        \item For every string $w$, $x$ and every natural number $k$,
            the predicate ``$\Phi(w, x) = k$?'' is decidable.
    \end{enumerate}
\end{definition}

@@

\begin{example}
    The standard measures of time and space can be constructed
    over the acceptable numbering of example~\ref{standard-numbering}.
    They are, respectively,
    the number of moves and tape cells scanned
    before halting.
    Leave the complexity undefined if the machine does not halt
    --- this satisfies the first axiom.
    The predicate of the second axiom is simple for time complexity;
    for space complexity,
    we must keep the whole history of computation
    to make sure the machine does not loop in a limited amount of space
    (because the complexity is not defined in this case).
\end{example}

\begin{definition}
    Given a complexity measure $\Phi$ for an acceptable Gödel numbering $\phi$
    and a total recursive function $f : \mathbb N \to \mathbb N$,
    define the \emph{complexity class} $\mathcal C_\Phi(f)$ by
    (\cite[p.~232]{Kozen2006})
    \begin{equation*}
        \mathcal C_\Phi(f) = \{
            \phi_w \mid \Phi(w, x) \leq f(|x|)\ \text{almost everywhere\footnotemark}
        \}.
    \end{equation*}
    \footnotetext{
        \emph{Almost everywhere} means that the inequality $\Phi_i(x) \leq f(|x|)$
        holds for all but a finite number of different $x$.
    }
\end{definition}

@@

Sections \ref{sec:godel-numberings} and~\ref{sec:blum-axioms}
introduced the concept of acceptable Gödel numbering,
the Blum axioms,
and its complexity classes.
As we want to regard these concepts as ``quality requirements''
for the definitions,
we will analyze each of them to see whether they define acceptable Gödel numberings,
the analogous time complexity defines a Blum complexity measure,
and that the characteristic function of the Boolean satisfiability problem
can be solved in ``polynomial time'' according to that complexity measure.

@@

\begin{proposition}
    The function partial $f: \Sigma^* \times \Sigma^* \to \{0, 1\}$ defined by
    \begin{equation*}
        f(w, x) = \begin{cases}
            1, & \text{if the machine $w$ does not halt on $x$.} \\
            \text{undefined}, &\text{if $w$ never halts on $x$.}
        \end{cases}
    \end{equation*}
    Can be computed by a nondeterministic Turing machine under
    Hopcroft and Ullman's definition.
\end{proposition}

@@

Therefore, with this restriction,
we lose the apparent exponential speed up in computation time
we have when using nondeterminism.

@@

So, Goldreich's definition defines an acceptable Gödel numbering
and we can form a complexity measure that satisfies Blum axioms.
But, again,
we have trouble with the ``exponential speedup'' requirement.
For instance,
a machine trying to solve the satisfiability problem
would correctly return $1$ for a satisfiable instance,
but no branch can write a $0$ alone
because it cannot be sure that instance is unsatisfiable
--- so, the function would be undefined for unsatisfiable formulas.

@@

\begin{proof}
    Suppose $M$ is the machine that computes $\SAT$'s characteristic function,
    under Goldreich definition, in polynomial time.

    The characteristic function of $\SAT$ is a total function,
    and its only outputs are $0$ and $1$.
    Therefore, in every computation of $M$,
    there is at least one branch that writes something different that $\bot$ on the tape,
    and whatever it writes is the correct answer.
    So, if we convert this to a nondeterministic decider and invert the output
    (branches that write $0$ will accept the input and vice-versa;
    branches that write $\bot$ always reject),
    any satisfiable formula will be rejected,
    because no branch of $M$ ever writes $0$ on the tape for these formulas,
    and any unsatisfiable formula will be accepted,
    because at least one branch of $M$ writes $0$ for these formulas.
    we can decide $\overline \SAT$, the complement of $\SAT$.

@@

Therefore, even though Goldreich's definition
yields an acceptable Gödel numbering and a Blum complexity measure,
it also have trouble in transposing the exponential speedup
we have using nondeterminism.

@@

\subsubsection{Extending $\NP$-completeness}
\label{sec:np-completeness-extension}

@@

\begin{definition}
    A function $f$ is $\FNP$-complete if $f \in \FNP$ and,
    for every function $g$ in $\FNP$,
    there is a polynomial-time computable function $h$ such that
    \begin{equation*}
        g(x) = f(h(x))
    \end{equation*}
    for every $x \in \{0, 1\}^*$.
\end{definition}

@@

\begin{definition}
    A function $f: \{0, 1\}^* \to \mathbb N$ is in $\OptP$
    if there is some nondeterministic Turing machine $M$ such that
    \cite[p.~493]{Krentel1988}
    \begin{itemize}
        \item For every input,
            every branch of $M$ halts within a polynomial number of steps
            and writes in its tape a number in binary; and
        \item Either,
            for all $x$, the largest number written by $M$ on $x$ is $f(x)$,
            or, for all $x$, the smallest number written by $M$ on $x$ is $f(x)$.
    \end{itemize}
\end{definition}

@@

\begin{definition}
    A function $f$ is $\OptP$-complete if $f \in \OptP$ and,
    for every function $g \in \OptP$,
    there are two functions $T_1 : \{0, 1\}^* \to \{0, 1\}^*$
    and $T_2 : \{0, 1\}^* \times \mathbb N \to \{0, 1\}^*$,
    both computable in deterministic polynomial time,
    such that
    \begin{equation*}
        g(x) = T_2( x, f(T_1(x)) ) \qquad \text{for all $x$.}
    \end{equation*}
\end{definition}

>>>>

In Theory of Computation,
we usually use languages
to mathematically model problems in the real world.
Decision problems (``yes/no'') are mapped to languages in a very natural way,
by just putting every ``yes'' instance in the language,
and leaving the rest out.
Search problems usually are rewritten as a decision problem,
and then this problem is converted to a language.
For instance,
the task of finding a satisfying assignment for a Boolean formula
is reinterpreted as the task of deciding whether such an assignment exists,
and this task is then converted to a language
--- in this example we have $\SAT$, the Boolean satisfiability problem.
This does the trick when it comes to proving that something is hard;
for instance, if we show that the decision problem is $\NP$-hard or undecidable,
then intuitively the corresponding search problem must be at least as hard.
Therefore,
concepts like ``decidable'', ``$\NP$-complete'', ``polynomial-time decidable''
arise naturally in the context of decision problems.

However, even from the theoretical standpoint,
it is useful to extend these concepts to functions.
For instance,
the concept of polynomial-time computable functions
are required to define Karp reductions \cite[p.~42]{AroraBarak2009}.
For single-tape deterministic Turing machines,
this definition is easy to extend:
A function $f : \{0, 1\}^* \to \{0, 1\}^*$
is said to be \emph{polynomial-time computable}
if there is a Turing machine $M$ and a polynomial $p$ such that,
when given the string $x$ as input,
$M$ halts with $f(x)$ in its tape within~$p(|x|)$~steps.

@@

This paper presents several attempts to establish
how a nondeterministic Turing machine could compute a function,
and to extend the concept of ``polynomial-time computable'' under each definition.
Section~\ref{sec:quality-standards} sets reasonableness criteria
to both the definition of function computation and complexity of the computation.
Section~\ref{sec:nondeterministic_computation_of_functions},
which contains the several definitions,
presents and criticizes Hopcroft and Ullman's and Goldreich's definitions
(sections \ref{sec:hopcroft-ullman} and~\ref{sec:goldreich}),
proposes one possible definition that meet the quality standards
(section~\ref{sec:proposed}),
and shows other two definitions,
by Krentel (sec~\ref{sec:krentel}) and Valiant (sec~\ref{sec:valiant}),
that, altough were not proposed in the context
of general nondeterministic computation of functions,
also meet the quality standards and sidesteps the problems with the proposed definition
(presented in section~\ref{sec:np-completeness-extension}).
The development of this project is mentioned in section~\ref{sec:development},
after the concluding remarks (section~\ref{sec:conclusion}).

@@

One important feature about the standard encodings of deterministic Turing machines
is the Universal Turing Machine Theorem:
the existence of a partial recursive function
$U: \{0, 1\}^* \times \{0, 1\}^* \to \{0, 1\}^*$
% Não sei se $U$ é um bom nome...
such that, if $w$ encodes the machine $M$,
then $U(m, x)$ is the result of running the machine $M$ on $x$.
A machines that compute $U$ is called \emph{universal Turing machine}.

@@

\vspace{6pt}
\begin{definition}
    Let $\mathcal P$ be the set of all partial recursive functions.
    An \emph{acceptable Gödel numbering}
    is a function $\phi : \{0, 1\}^* \to \mathcal P$,
    that associates each string (or program\footnotemark) $w \in \{0, 1\}^*$
    \footnotetext{
        In texts like Rogers'~\cite{Rogers1987},
        the partial recursive functions have the naturals as domain and codomain
        (that is, they are of the form $f: \mathbb N \to \mathbb N$),
        and Gödel numberings associates natural numbers with recursive functions.
        In this paper we will work with binary strings instead of numbers,
        which will simplify the definition of complexity classes
        and allows us to think the string $w$ as a \emph{program} for $\phi_w$
        (see example~\ref{c}).
    }
    to a function $\phi_w \in \mathcal P$,
    that satisfies
    \begin{enumerate}
        \item $\phi$ is surjective;
            that is, every partial recursive function $f \in \mathcal P$
            has a program $w \in \{0, 1\}^*$ such that $\phi_w = f$;
            \label{surjectiveness}
        \item There is a partial recursive function
            $U:\{0, 1\}^* \times \{0, 1\}^* \to \{0, 1\}^*$ such that,
            for every $w$ and $x$,
            $U(w, x)$ is defined if and only if $\phi_w(x)$ is defined,
            and, in this case,
            \begin{equation*}
                U(w, x) = \phi_w(x);
            \end{equation*}
            \label{universal-tm}
            \vspace{-18pt}
        \item There is a total recursive function
            $\sigma:\{0, 1\}^* \times \{0, 1\}^* \to \{0, 1\}^*$ such that,
            for every $y$,
            $\phi_w(x, y)$ is defined if and only if $\phi_{\sigma(w, x)}(y)$ is defined,
            and, in this case,
            \begin{equation*}
                \phi_w(x, y) = \phi_{\sigma(w, x)}(y).
            \end{equation*}
            \label{smn-theorem}
            \vspace{-18pt}
    \end{enumerate}
\end{definition}

@@

Condition~\ref{smn-theorem} is the ``little more'' we mentioned earlier.
It is known as the $S_{mn}$ theorem \cite[p.~24]{Rogers1987}.
Essentially,
given a partial recursive function of two variables,
we can obtain a partial recursive function of one variable
by fixing the first argument.
The function $\sigma$ provides a systematic way of doing this:
given a description $w$ of a partial recursive function of two variables
and the value $x$ to be fixed as the first variable,
$\sigma(w, x)$ is a machine that computes this new partial recursive function.

\vspace{6pt}
\begin{example}\label{standard-numbering}
    Any encoding of deterministic Turing machines as a binary string
    yields an acceptable numbering of the recursive functions.
\end{example}

\vspace{6pt}
\begin{example}\label{c}
    Any programming language can be understood as an acceptable Gödel numbering.
    For example,
    if we restrict a \texttt C program
    to perform input and output only using the standard input and standard output
    (that is, forbid interactions with the user, file reading, GUIs,
    access to system clock, etc.),
    the resulting program will map a binary input to a binary output,
    characterizing a partial recursive function.
    Thus, we can see a \texttt C compiler
    as an implementation of an acceptable Gödel numbering.\footnote{
        Note we are ignoring here issues like compilation and run-time errors.
        These can be dealt with as in the case of Turing machines:
        any invalid program will signify some fixed partial recursive function
        (say, the function that is defined nowhere);
        and any invalid step in computation
        makes the function to be not defined on that input.
    }
\end{example}

@@

\vspace{12pt}
\begin{definition}
    Given an acceptable Gödel numbering $\phi$,
    a \emph{complexity measure} for $\phi$
    is a function $\Phi:\Sigma^* \times \Sigma^* \to \mathbb N$ of two variables
    that satisfies \cite[p.~324]{Blum1967}:
    \begin{enumerate}
        \item For every $w$ and $x$,
            $\phi_w(x)$ exists if and only if $\Phi(w, x)$ exists; and
        \item For every string $w$, $x$ and every natural number $k$,
            the predicate ``$\Phi(w, x) = k$?'' is decidable.
    \end{enumerate}
\end{definition}

@@

\vspace{12pt}
\begin{example}
    The standard measures of time and space can be constructed
    over the acceptable numbering of example~\ref{standard-numbering}.
    They are, respectively,
    the number of moves and tape cells scanned
    before halting.
    Leave the complexity undefined if the machine does not halt
    --- this satisfies the first axiom.
    The predicate of the second axiom is simple for time complexity;
    for space complexity,
    we must keep the whole history of computation
    to make sure the machine does not loop in a limited amount of space
    (because the complexity is not defined in this case).
\end{example}

\vspace{12pt}
\begin{definition}
    Given a complexity measure $\Phi$ for an acceptable Gödel numbering $\phi$
    and a total recursive function $f : \mathbb N \to \mathbb N$,
    define the \emph{complexity class} $\mathcal C_\Phi(f)$ by
    (\cite[p.~232]{Kozen2006})
    \begin{equation*}
        \mathcal C_\Phi(f) = \{
            \phi_w \mid \Phi(w, x) \leq f(|x|)\ \text{almost everywhere\footnotemark}
        \}.
    \end{equation*}
    \footnotetext{
        \emph{Almost everywhere} means that the inequality $\Phi_i(x) \leq f(|x|)$
        holds for all but a finite number of different $x$.
    }
\end{definition}

@@

Sections \ref{sec:godel-numberings} and~\ref{sec:blum-axioms}
introduced the concept of acceptable Gödel numbering,
the Blum axioms,
and its complexity classes.
As we want to regard these concepts as ``quality requirements''
for the definitions,
we will analyze each of them to see whether they define acceptable Gödel numberings,
the analogous time complexity satisfies Blum axioms,
and that the characteristic function of the Boolean satisfiability problem
can be solved in ``polynomial time'' according to that complexity measure.

@@

\vspace{6pt}
\begin{proposition}
    Define the partial function $f: \Sigma^* \times \Sigma^* \to \{0, 1\}$ by
    \begin{equation*}
        f(w, x) = \begin{cases}
            1, & \text{if the machine $w$ does not halt on $x$.} \\
            \text{undefined}, &\text{if $w$ never halts on $x$.}
        \end{cases}
    \end{equation*}
    This function can be computed by a nondeterministic Turing machine under
    Hopcroft and Ullman's definition.
\end{proposition}
\vspace{-3pt}

@@

So, with this restriction,
we lose the apparent exponential speed-up in computation time
we have when using nondeterminism.

@@

So, Goldreich's definition defines an acceptable Gödel numbering
and we can form a complexity measure that satisfies Blum axioms.
But, again,
we have trouble with the ``exponential speed-up'' requirement.
For instance,
a machine trying to solve the satisfiability problem
would correctly return $1$ for a satisfiable instance,
but no branch can write a $0$ alone
because it cannot be sure that instance is unsatisfiable
--- so, the function would be undefined for unsatisfiable formulas.

@@

\begin{proof}
    Suppose $M$ is the machine that computes $\SAT$'s characteristic function,
    under Goldreich's definition, in polynomial time.

    The characteristic function of $\SAT$ is a total function,
    and its only outputs are $0$ and $1$.
    Therefore, in every computation of $M$,
    there is at least one branch that writes something different that $\bot$ on the tape,
    and whatever it writes is the correct answer.
    So, if we convert this to a nondeterministic decider and invert the output
    (branches that write $0$ will accept the input and vice-versa;
    branches that write $\bot$ always reject),
    any satisfiable formula will be rejected,
    because no branch of $M$ ever writes $0$ on the tape for these formulas;
    and any unsatisfiable formula will be accepted,
    because at least one branch of $M$ writes $0$ for these formulas.
    Thus, we can decide $\overline \SAT$, the complement of $\SAT$.

@@

Therefore, even though Goldreich's definition
yields an acceptable Gödel numbering and a Blum complexity measure,
it also have trouble in transposing the exponential speed-up
we have using nondeterminism.

@@

\vspace{-12pt}
\subsubsection{Extending $\NP$-completeness}
\label{sec:np-completeness-extension}

@@

\vspace{6pt}
\begin{definition}
    A function $f$ is $\FNP$-complete if $f \in \FNP$ and,
    for every function $g$ in $\FNP$,
    there is a polynomial-time computable function $h$ such that
    \begin{equation*}
        g(x) = f(h(x))
    \end{equation*}
    for every $x \in \{0, 1\}^*$.
\end{definition}

@@

\begin{definition}
    A function $f: \{0, 1\}^* \to \mathbb N$ is in $\OptP$
    if there is some nondeterministic Turing machine $M$ such that
    \cite[p.~493]{Krentel1988}:
    \begin{itemize}
        \item For every input,
            every branch of $M$ halts within a polynomial number of steps
            and writes in its tape a number in binary; and
        \item Either,
            for all $x$, the largest number written by $M$ on $x$ is $f(x)$,
            or, for all $x$, the smallest number written by $M$ on $x$ is $f(x)$.
    \end{itemize}
\end{definition}

@@

\vspace{6pt}
\begin{definition}
    A function $f$ is $\OptP$-complete if $f \in \OptP$ and,
    for every function $g \in \OptP$,
    there are two functions $T_1 : \{0, 1\}^* \to \{0, 1\}^*$
    and $T_2 : \{0, 1\}^* \times \mathbb N \to \{0, 1\}^*$,
    both computable in deterministic polynomial time,
    such that, for all $x$,
    \begin{equation*}
        g(x) = T_2\big( x, f(T_1(x)) \big).
    \end{equation*}
\end{definition}
